---
title: "5 Working With Numbers"
author: "Gustavo R. Santos (original) | Antti Rask (modifications)"
date: "2022-07-11"
output: html_document
---

# Creating Numeric Variables

## Load Libraries

```{r, message=FALSE, warning=FALSE}
library(conflicted) # An Alternative Conflict Resolution Strategy
  conflicts_prefer(dplyr::filter)
library(patchwork)  # The Composer of Plots
library(skimr)      # Compact and Flexible Summaries of Data
library(tidyverse)  # Easily Install and Load the 'Tidyverse'
```

## Primer

```{r}
# Creating a scalar
scalar <- 42
scalar

# Creating a vector
vector <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
vector

# Creating a Matrix
matrix <- vector %>%
  matrix(nrow = 3, ncol = 3)

matrix
```

```{r}
# Creating a tibble
tbl <- tibble(
  column1 = c(1,2,3),
  column2 = c(4,5,6),
  column3 = c(7,8,9)
)

tbl
```

## Vectors

```{r}
# Vector
vec1 <- c(9,8,7,6,5,4,3,2,1)
vec2 <- c(1,2,3,4,5,6,7,8,9)

# Slicing a vector
vec1[1:3]

# Slice by condition
vec1[vec1 >= 6]

# Sum of vectors
vec1 + vec2

# Multiplication of vectors
vec1 * vec2

# Addition or Multiplication of a vector by a single number
vec1 * 10
vec1 - 1

# Two vectors of different sizes
c(1, 2) + c(1, 2, 3, 4)
```

## Matrix

```{r}
# Create a Matrix
matrix <- matrix(
  1:12,
  nrow  = 4,
  ncol  = 3,
  byrow = TRUE
)

matrix
```

```{r}
# Slicing [row,col]
matrix[2, 3]
```

## Tibbles

```{r}
# Tibble
tbl <- tibble(
  name   = c("Carl", "Vanessa", "Hanna", "Barbara"),
  class_ = c("Math", "Math", "Math", "Math"),
  grade  = c(8.5, 9, 9, 7)
)

tbl
```

```{r}
# Slicing Carl's grade (col 1 and col 3)
tbl %>%
  slice(1) %>%
  pull(grade)

# Alternatively
tbl %>%
  filter(name == "Carl") %>% 
  pull(grade)
```

```{r}
# Slicing by variable
tbl %>% 
  pull(grade)
```

```{r}
# Slicing by condition
tbl %>% 
  filter(grade > 8)
```

### Example: Extra Point from the Professor

```{r}
tbl_with_new_grade <- tbl %>% 
  mutate(new_grade = grade + 1)

tbl_with_new_grade
```

### Example: Normalization

```{r}
max_grade <- tbl_with_new_grade %>% 
  pull(grade) %>% 
  max()

tbl_normalized <- tbl_with_new_grade %>% 
  mutate(normalized_grade = grade / max_grade)

tbl_normalized
```

## Map Functions from {purrr} and _Summarize()_ to Replace the _apply()_ Family of Functions

### pmap_dbl()

```{r}
# apply function mean to rows: sum 'grade' + 'new_grade'
tbl_normalized %>% 
  select(grade, new_grade) %>% 
  pmap_dbl(~ mean(c(...)))
```

```{r}
# apply function mean to columns the column 3 and 4
tbl_normalized %>% 
  select(grade, new_grade) %>% 
  summarise(across(everything(), mean))
```

```{r}
# apply custom function
calculate_grade_new_grade_avg <- function(grade, new_grade) {
  sum(grade, new_grade) / 2
}

tbl_normalized %>% 
  select(grade, new_grade) %>% 
  pmap_dbl(calculate_grade_new_grade_avg)
```
 
### map()

#### Remove the Extra Point from new_grade

```{r}
remove_a_point <- function (x){x - 1}
map_obj <- map(tbl_normalized %>% pull(new_grade), remove_a_point)
map_obj
```

### map_dbl()

```{r}
# map_dbl a custom function to a single column
# my_func <- function(x) {x - 1}
tbl_normalized %>%
  pull(new_grade) %>% 
  map_dbl(remove_a_point)

# mapping of a custom function to two columns
tbl_normalized %>%
  select(grade, new_grade) %>% 
  map(remove_a_point) %>%
  as_tibble() %>% 
  as.matrix()
```

### Summarize

```{r}
# Create new tibble
tbl <- tibble(
  name   = c("Carl", "Vanessa", "Hanna", "Barbara", "Jason", "Alison", "Kevin", "Melody"),
  class_ = c("Math", "Math", "Math", "Math", "Math2", "Math2", "Math2", "Math2"),
  grade  = c(8.5, 9, 9, 7, 5, 7, 10, 9.5)
)

# Calculate the mean for each class_
tbl %>% 
  summarise(
    mean_grade = mean(grade),
    .by        = class_
  )
```
 
## Descriptive Statistics

```{r}
# Set a seed to reproduce the same results
set.seed(12)

# Creating a variable: 50 data points, mean 0, standard deviation 1
var1 <- rnorm(50, mean = 0, sd = 1)

# Creating a variable: 50 data points, mean 0, standard deviation 10
var2 <- rnorm(50, mean = 0, sd = 5)

# Mean
mean(var1)
mean(var2)
```

```{r}
# Standard Deviation
sd(var1)
sd(var2)
```

```{r}
# Median
median(var1)
median(var2)
```

```{r}
# percentiles or quantiles
quantile(var1, c(0.25, 0.5, 0.75))
quantile(var2, c(0.25, 0.5, 0.75))
```
 
### Histograms

```{r}
# Histogram var1
p1 <- var1 %>% 
  as_tibble() %>% 
  ggplot(aes(var1)) +
  geom_histogram(
    breaks = seq(-2, 3, by = 0.5),
    fill   = "royalblue",
    color  = "black"
  ) +
  scale_x_continuous(
    limits = c(-20, 20) 
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0, 20)
  ) +
  labs(y = "Frequency")

# Histogram var2
p2 <- var2 %>%
  as_tibble() %>% 
  ggplot(aes(var2)) +
  geom_histogram(
    breaks = seq(-15, 15, by = 5),
    fill   = "coral",
    color  = "black"
  ) +
  scale_x_continuous(
    limits = c(-20, 20) 
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0, 20)
  ) +
  labs(y = "Frequency")
  
(p1 + p2) * theme_classic()
```
 
### Descriptive Statistics in Numbers

```{r}
# Create a tibble with both variables for comparison
tbl <- tibble(
  var1 = var1,
  var2 = var2
)

# Comparison tibble

# Define a function to apply summary statistics
apply_summary_stats <- function(x) {
  list(
    avg        = mean(x),
		std_dev    = sd(x),
		min_val    = min(x),
		pct25      = quantile(x, 0.25),
		median_val = median(x),
		pct75      = quantile(x, 0.75),
		max_val    = max(x)
  )
}

# Use the function with the pipeline
tbl %>%
	reframe(
		var   = names(.),
		stats = map(var, ~ apply_summary_stats(tbl[[.x]]))
	) %>%
	unnest_wider(stats) %>%
	pivot_longer(
		cols      = -var,
		names_to  = "stat",
		values_to = "value"
	) %>%
	pivot_wider(
		names_from  = "stat",
		values_from = "value"
	)
```

### Calculate Statistics Using _summary()_ and/or _skim()_

```{r}
# Descriptive stats with summary function
summary(tbl)
```

```{r}
skim(tbl)
```

### Correlation

```{r}
cor(var1, var2, method = "pearson")
```

#### Scatterplot

```{r}
tbl %>% 
  ggplot(aes(var1, var2)) +
  geom_point(color = "royalblue", size = 2) +
  labs(title = "Scatterplot var1 x var2") +
  theme_classic()
```
