---
title: "7 Transformations with base R"
author: "Gustavo R. Santos (original) | Antti Rask (modifications)"
date: "2022-07-21"
output: html_document
---

# Transformations with base R

Antti: This whole time I've been transforming base R code to {tidyverse}. And I will continue doing it for the other chapters. This one is the exception, though. Even so, I will still use things like tibble and the pipe operator when it makes sense to do so.

## Load Libraries

```{r, message=FALSE, warning=FALSE}
library(conflicted) # An Alternative Conflict Resolution Strategy 
library(data.table) # Extension of `data.frame` 
library(skimr)      # Compact and Flexible Summaries of Data
library(tidyverse)  # Easily Install and Load the 'Tidyverse'
```

## Dataset

The dataset to be used in this chapter is the _Adult Data Set_ from the UCI Machine Learning Repository.

__Dataset Credits:__
Dua, D. and Graff, C. (2019). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science.

__URL Address__
https://archive.ics.uci.edu/ml/datasets/Adult

### Loading the dataset

```{r}
# Load the dataset to RStudio
tbl <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data",
  header      = FALSE,
  strip.white = TRUE
) %>%
  as_tibble()

# Add column names
colnames(tbl) <- c(
  "age",
  "workclass",
  "fnlwgt",
  "education",
  "education_num",
  "marital_status",
  "occupation",
  "relationship",
  "race",
  "sex",
  "capital_gain",
  "capital_loss",
  "hours_per_week",
  "native_country",
  "target"
)

head(tbl)
```

### Slicing

Slicing is like zooming in our dataset, looking only the parts we need or want to see.

```{r}
# Dataset dimensions
dim(tbl)
```

```{r}
# Slicing rows 1 to 3, all columns
tbl[1:3, ]
```

```{r}
# Slicing all rows, columns 1 to 3
tbl[, 1:3]
```

```{r}
# Slicing with conditional, vector for cols 1 and 6
tbl[tbl$age > 30, c(1,6)]
```

### Filtering

Filter differs from slice because the filter returns all the variables, so it won't be able to cut out some variables.

```{r}
# Filter
tbl %>%
  subset(age > 30)
```

```{r}
# Slice with subset, returning only the selected columns
tbl %>% 
  subset(age > 30, select = c(age, marital_status))
```

### Replace and Fill values

#### Replace

Replace is to exchange a value for another.
Fill values is when you have a _NA_ value that needs to be filled with an average or another value, like zero.

```{r}
# Check for NA values
sum(is.na(tbl))
```

```{r}
# Replace "?" using slicing
tbl$occupation[tbl$occupation == "?"]            <- NA
tbl[tbl$native_country == "?", "native_country"] <- NA

# Replace "?" with NA using gsub
tbl$workclass <- gsub(
  pattern     = "\\?",
  replacement = NA, 
  tbl$workclass
)

# Replace "?" with ifelse
tbl$native_country <- ifelse(tbl$native_country == "?", NA, tbl$native_country)

# Check for NA values
sum(is.na(tbl))
```

#### Check missing data by column

```{r}
skim(tbl)
```

#### Filling

Common fill values are zero, mean, median, most frequent (for categorical).
There are more complex ways to fill, using machine learning, but that is not in the scope of this book.

```{r}
# Fill NAs with most frequent value
most_frequent <- names(table(tbl$workclass)[which.max(table(tbl$workclass))])
tbl$workclass[is.na(tbl$workclass)] <- most_frequent

# Occupation
most_frequent <- names(table(tbl$occupation)[which.max(table(tbl$occupation))])
tbl$occupation[is.na(tbl$occupation)] <- most_frequent

# Remove all NAs (the remaining in native_country)
tbl <- na.omit(tbl)

tbl
```

Filling NAs from Numeric columns

```{r}
# Data frame
tbl_num <- tibble(
  A = c(1, 2, 2, 2, 3, NA),
  B = c(3, 4, 5, 3, NA, 0),
  C = c(1, 1, 1, NA, NA, 5)
)

# Fill NAs with 0
nafill(tbl_num, type = "const", fill = 0)

# Fill NAs with the last valid number
nafill(tbl_num, type = "locf")

# Fill NAs with the next valid number
nafill(tbl_num, type = "nocb")

# Fill NAs with mean
tbl_num$A[is.na(tbl_num$A)] <- mean(tbl_num$A, na.rm = TRUE)
```

```{r}
tbl_num
```

### Arranging Data

Arrange data is useful to rank items from top to bottom or vice-versa, creating ordinated datasets.

```{r}
# Arrange raw data
tbl[order(tbl$age, decreasing = TRUE), ]

# Arrange raw data using - instead of "decreasing = TRUE"
tbl[order(-tbl$age), ]
```

### Creating New Variables

During data wrangling, it is common to create new variables, being that for splitting an information or combining two or more variables to make a new one.

Note: to create new columns in R, you can use _tbl$new_col_name_ or _tbl['new_col_name]_

```{r}
# Binarizing sex using ifelse()
tbl$sex_binary <- ifelse(tbl$sex == "Female", 1, 0)
tbl[ , c(10, 16)]
```

To Encode a variable, transforming the names in numbers, here is a solution.
We will:

1. Assign the variable _education_ as a factor
2. Create order to the levels
3. Store those orders in another variable
4. Change the new variable to numbers

```{r}
## Encoding a variable ##

# Assign education as factor
tbl$education <- as.factor(tbl$education)

# Order the factors
tbl$education <- ordered(
  tbl$education,
  levels = c(
    "Preschool",
    "1st-4th",
    "5th-6th",
    "7th-8th",
    "9th",
    "10th",
    "11th",
    "12th",
    "HS-grad",
    "Some-college",
    "Assoc-acdm",
    "Assoc-voc",
    "Bachelors",
    "Masters",
    "Doctorate"
  )
)

# Create new variable
tbl$education_cd <- tbl$education

# Change levels to numbers
levels(tbl$education_cd) <- 1:15
tbl[, c("education", "education_cd")]
```

#### Creating a custom calculation

Imagine that there were a rule in place that, for every _capital_gain_ - _capital_loss_ equal or greater than 15,000 there was a 10% tax.

We will first create a _total_ variable to show arithmetic operation and then calculate the 10% tax

```{r}
# Total gain variable creation
tbl$total_gain <- tbl$capital_gain - tbl$capital_loss

# Tax variable creation
tbl$tax <- ifelse(tbl$total_gain >= 15000, tbl$total_gain * 0.1, 0)

tbl[ , c(11,12,17,18)][order(-tbl$tax), ]
```

```{r}
# Shorter code

# Tax variable creation
tbl$tax2 <- ifelse(tbl$total_gain >= 15000, (tbl$capital_gain - tbl$capital_loss) * 0.1, 0)

tbl
```

### Binding

Binding data is useful to glue pieces of data together.

_rbind()_ or bind rows glues the data by columns. So if your data has the same variables, you can use this function.

```{r}
# Creating datasets A and B
A <- tbl[1:10, ]
B <- tbl[11:20, ]

# Append / bind rows
AB <- rbind(A, B)
AB
```

_cbind()_ is to bind columns, gathering the data by rows. It is more suitable when the rows are for the same observation, but the columns are different and need to be put together.

```{r}
# Creating datasets A and B
A <- tbl[1:10, 1:3]
B <- tbl[1:10, c(4, 7)]

# Bind columns
AB <- cbind(A, B)
AB
```

## data.table

Review of the library _data.table_

Basic syntax

DT[i, j, by]
* *i* is for the rows selection or a condition for the rows to be displayed
* *j* is to select the variables or to calculate a statistic of it
* *by* is used to group variables

Let's start by recreating the dataset:

```{r}
# Load the dataset to RStudio
tbl <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data",
  header      = FALSE,
  strip.white = TRUE
) %>%
  as_tibble()

# Add column names
colnames(tbl) <- c(
  "age",
  "workclass",
  "fnlwgt",
  "education",
  "education_num",
  "marital_status",
  "occupation",
  "relationship",
  "race",
  "sex",
  "capital_gain",
  "capital_loss",
  "hours_per_week",
  "native_country",
  "target"
)

head(tbl)
```

```{r}
# Assign variable as data.table object
dt <- as.data.table(tbl)

dt
```

```{r}
# Syntax
dt[dt$age > 50, .(age, occupation)]
```

```{r}
# Mean age Group By relationship
dt[, mean(age), by = relationship][order(-V1),]
```

### Group by and summarize

Group by will put the data in groups. Summarize will reduce the aggregated data values to one number, like mean, median, count etc.

```{r}
# Group By and count number of observations
dt[, .N, by = sex]
```

```{r}
# Group By and return mean of age by sex
dt[, mean(age), by = sex]
```

```{r}
# Group By and return sum of capital_gain by sex
dt[, sum(capital_gain), by = sex]
```

### Group and visualize

```{r}
# Mean education years by education level
educ_yrs <- dt[, mean(education_num), by = education]
educ_yrs <- educ_yrs[order(V1)]

# barplot
barplot(
  educ_yrs$V1,
  names.arg = educ_yrs$education,
  col       = "royalblue",
  main      = "Average years of education by grade level"
)
```

### Arranging data

Using data.table, we can group variables and order them.

```{r}
# Group and order education years average by workclass
dt[, mean(education_num), by = workclass][order(-V1),]
```
