  ---
title: "8 Transformations with tidyverse"
author: "Gustavo R. Santos (original) | Antti Rask (modifications)"
date: "2022-07-26"
output: html_document
---

# Transformations with Tidyverse Libraries

## Load Libraries

```{r, message=FALSE, warning=FALSE}
library(conflicted) # An Alternative Conflict Resolution Strategy
  conflicts_prefer(dplyr::filter)
library(tidyverse)  # Easily Install and Load the 'Tidyverse'
```

## Dataset

The dataset to be used in this chapter is the _Adult Data Set_ (also know as _Census Income_) from the UCI Machine Learning Repository.

__Dataset Credits:__
Dua, D. and Graff, C. (2019). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science.

__URL Address__
https://archive.ics.uci.edu/ml/datasets/Adult

### Loading the dataset

```{r}
# Define column names
header <- c(
  "age",
  "workclass",
  "fnlwgt",
  "education",
  "education_num",
  "marital_status",
  "occupation",
  "relationship",
  "race",
  "sex",
  "capital_gain",
  "capital_loss",
  "hours_per_week",
  "native_country",
  "target"
)

# Load the dataset to RStudio
tbl <- read_csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data",
  col_names = header,
  trim_ws   = TRUE
)
```

```{r}
head(tbl)
```

### Slicing

Slicing is like zooming in our dataset, looking only the parts we need or want to see.

```{r}
# Slicing rows 1 to 5, columns 1 to 4.
tbl %>%
  slice_head(n = 5) %>% 
  select(1:4)
```

```{r}
# Slicing with slice_min() and slice_max()
tbl %>% 
  slice_min(age, prop = 0.10)

tbl %>%
  slice_max(age, prop = 0.30)
```

```{r}
# Slice sample
tbl %>%
  slice_sample(
    n       = 10,
    replace = TRUE
  )
```

### Filtering

Filter returns all the variables for the observations that fulfill the filter condition. Using _select()_, we can also select only the wanted variables.

```{r}
# Filtering age over 30 years old
tbl %>%
  filter(age > 30)
```

```{r}
# Filter age > 30 and selecting age and marital_status
tbl %>% 
  filter(age > 30) %>% 
  select(marital_status, age)
```

```{r}
# Distinct - removing duplicates
tbl %>%
  distinct(sex)
```

### Group by and summarize

Group by will put the data in groups. Summarize will reduce the aggregated data values to one number, like mean, median, count etc.

This is how the functions behave when called alone

```{r}
# group by not summarized
tbl %>%
  group_by(workclass)
```

```{r}
# summarise without group_by returning average age of the dataset
tbl %>%
  summarise(age_avg = mean(age))
```

```{r}
# Group By workclass and Summarize mean age -> .by
tbl %>% 
  summarise(
    age_avg = mean(age),
    .by     = workclass
  )
```

Let's look at the object returned by dataset grouped by more than one variable. Notice that it is a grouped tibble. To make the return object be a regular tibble, use the function _ungroup()_.

```{r}
# Returns object grouped_tbl
workclass_groups <- tbl %>%
  group_by(workclass, sex) %>% 
  summarise(age_avg = mean(age))

workclass_groups
```

```{r}
# Returns object tibble
workclass_ungrouped <- tbl %>%
  summarise(
    age_avg = mean(age),
    .by     = c(workclass, sex)
  )

workclass_ungrouped
```

#### Summary Functions

Functions to be used with _summarise()_.

```{r}
# n() shows the count of observations in each group
tbl %>%
  summarise(
    n(),
    .by = workclass
  )
```

```{r}
# n_distinct() shows the count of unique observations in each group
tbl %>%
  summarize(n_distinct(workclass))

# if you want to dig deeper
tbl %>% 
  summarize(n_distinct(native_country), .by = workclass)
```

```{r}
# sum(!is.na()) shows the count of Non NA observations
tbl %>% summarise(sum(!is.na(workclass)))
```

```{r}
# first() shows the first age value in each group
# Similarly, you can use last() or nth()
tbl %>%
  summarise(
    first(age),
    .by = workclass
  )
```

```{r}
# quantile() shows the top number that meets the quantile percentage chosen
# In this case, 50% of the age observations are under what value by group
tbl %>% summarise(
  quantile(age, 0.5),
  .by = workclass
)
```

```{r}
# sd() shows standard deviation of a variable
tbl %>%
  summarise(
    sd(capital_gain),
    .by = workclass
  )
```

```{r}
# Across function, to apply the function mean to the selected columns
tbl %>%
  select(1, 3, 5, 11, 12, 13) %>%
  summarise(across(everything(), mean))
```

### Replace and Fill values

For this part of the exercise, let's create a custom function to count the NAs by column. It will be useful during the exercise.


```{r}
na_by_column <- function(tbl) {
  
  # Function that takes in a data frame and returns a table with the NA count by column.
  # Input:
  # tbl: data frame
  # Return
  # tibble object
  
  tbl %>%
    summarise(across(everything(), ~sum(is.na(.x)))) %>%
    pivot_longer(
      cols      = everything(),
      names_to  = "Column",
      values_to = "NA count"
    ) %>%
    arrange(Column)
}
```

#### Replace

Replace is to exchange a value for another. Here, we will replace the `?` with `NA`, so we can fill the NAs later.

The first way to do that is using replace. 

```{r}
# Define a function to count '?' values in each column
count_question_marks <- function(x) {
  sum(x == "?", na.rm = TRUE)
}
```

```{r}
# Print the count of '?' values in each column
tbl %>%
  summarise(across(everything(), count_question_marks)) %>%
  pivot_longer(
    cols      = everything(),
    names_to  = "Column",
    values_to = "Count of '?'"
  ) %>%
  filter(`Count of '?'` > 0)
```

```{r}
# Replace '?' values with NA in a new dataset variable
tbl_replaced <- tbl %>%
  mutate(across(where(is.character), ~replace(., . == "?", NA)))

# Count NAs by variable using the previously defined na_by_column() function
na_by_column(tbl_replaced)
```

#### NA_IF

But there is another way, much easier, by the way, using the function _na_if()_. Simply put, if a value is a given pattern, make it _NA_.

```{r}
# Replacing values "?" with NA (tidyverse 1.3.2)
tbl_replaced <- tbl %>%
  mutate(
    workclass      = na_if(workclass, "?"),
    occupation     = na_if(occupation, "?"),
    native_country = na_if(native_country, "?")
  )

# Count NAs by variable
na_by_column(tbl_replaced)
```

#### Filling

Fill values is when you have a _NA_ value that needs to be filled with an average or another value, like zero.

```{r}
# Fill NA values with last or next valid value
tbl_replaced %>% fill(
  workclass,
  occupation,
  native_country,
  .direction = "down"
)
```

```{r}
# Finding the most frequent entry
get_most_frequent <- function(data, column) {
  data %>%
    summarise(
      freq = n(),
      .by  = all_of(column)
    ) %>%
    top_n(1, freq) %>%
    pull(column)
}

m_freq_workcls <- get_most_frequent(tbl, "workclass")
m_freq_workcls

m_freq_occup   <- get_most_frequent(tbl, "occupation")
m_freq_occup
```

```{r}
# Replace NA with the most frequent value
tbl_no_na <- tbl_replaced %>%
  replace_na(list(workclass = m_freq_workcls, occupation = m_freq_occup)) %>% 
  # Drop NAs and save the dataset in a new variable name
  drop_na()

tbl_no_na
```

#### Filling NAs from Numeric columns

```{r}
# Tibble
tbl_num <- tibble(
  A = c(1, 2, 2, 2, 3, NA),
  B = c(3, 4, 5, 3, NA, 0),
  C = c(1, 1, 1, NA, NA, 5)
)

# Fill NAs for numeric columns with statistics measurements
tbl_num %>% 
  replace_na(
    list(
      A = median(tbl_num$A, na.rm = TRUE),
      B = mean(tbl_num$B, na.rm = TRUE),
      C = min(tbl_num$C, na.rm = TRUE)
    )
  )
```

### Arranging Data

Arrange data is useful to rank items from top to bottom or vice-versa, creating ordinated datasets.

```{r}
# Arrange data in increasing order
tbl_no_na %>% arrange(native_country)
```

```{r}
# Arrange data in decreasing order
tbl_no_na %>% arrange(desc(native_country))
```

Using _tidyverse_ to group variables and order them.

```{r}
# Group and order average net gain by education level
tbl_no_na %>%
  summarise(
    count        = n(),
    avg_net_gain = mean(capital_gain - capital_loss),
    .by          = education
  ) %>% 
  arrange(desc(avg_net_gain))
```

### Creating New Variables

During data wrangling, it is common to create new variables, being that for splitting an information or combining two or more variables to make a new one.

__Note: to create new columns in R, you can use `tbl$new_col_name` or `tbl['new_col_name]`__

```{r}
# Since the delimiter used was the word boundary, we couldn't replace the superseded separate() function with spread_wider_delim()

# Split variable target into sign and amount
tbl_no_na %>%
  separate_wider_regex(
    target,
    c(
      amt  = "\\D*(?<=\\b)",
      sign = ".*"
    )
  )
```

```{r}
# Unite variables sex, race and age
tbl_no_na %>% 
  unite(
    sex,
    race,
    age,
    col    = "description",
    sep    = "_",
    remove = FALSE
  )
```

#### Creating a custom calculation

Imagine that there were a rule in place that, for every _capital_gain_ - _capital_loss_ equal or greater than 15,000 there was a 10% tax.

We will first create a _total_ variable to show arithmetic operation and then calculate the 10% tax

```{r}
# Tax variable creation
tbl_no_na %>% 
  mutate(
    total_gain = capital_gain - capital_loss,
    tax        = case_when(
      total_gain >= 15000 ~ total_gain * 0.1,
                     TRUE ~ 0
    )  
  ) %>% 
  arrange(desc(tax))
```

Let’s change <=50K to _under_ and >50K to _over_.

```{r}
# Change values from target to over or under
 tbl_no_na %>% 
  mutate(
    over_under = case_match(
      target,
      "<=50K" ~ "under",
      ">50K"  ~ "over"
    )
  ) %>%
  select(target, over_under)
```

Mutate and cut to see which observations are over or under the average.

```{r}
# Observations over or under the average
tbl_no_na %>% 
  mutate(
    age_avg            = mean(age, na.rm = TRUE),
    over_under_age_avg = case_when(
      age <  mean(age, na.rm = TRUE) ~ "Lower than avg",
      age >= mean(age, na.rm = TRUE) ~ "Above the avg"
    )
  ) %>%
  select(age, age_avg, over_under_age_avg)
```

### Reshape

Reshaping means to transform the dataset from wide format to long or vice versa.

Roughly, we can say that:
Wide format: variables are rows.
Long format: variables are columns.

#### Wide to Long

```{r}
# set seed to generate the same random numbers every run
set.seed(42)

# Create a dataset
tbl_wide <- tibble(
  project = c("project1", "project2", "project3"),
  Jan = sample(1000:2000, 3),
  Feb = sample(1000:2000, 3),
  Mar = sample(1000:2000, 3)
)
  
# Wide to Long
tbl_long <- tbl_wide %>% 
  pivot_longer(
    cols      = Jan:Mar,
    names_to  = "months",
    values_to = "expenses"
  )

tbl_long
```

#### Long to Wide

```{r}
# Long to Wide
tbl_long %>%
  pivot_wider(
    names_from  = "months",
    values_from = "expenses"
  )
```

# Joining Datasets

## Creating tables

Suppose you are a data scientist or engineer in a retail company and have three tables. One fact table, with measurements, and two dimension tables, with descriptions. Our goal is to use tidyverse functions to create all types of joins with the datasets.

```{r}
# Fact table
sales <- tibble(
  date       = c("2022-01-01", "2022-01-02", "2022-01-03", "2022-01-04", "2022-01-05"),
  store_cd   = c(1, 2, 3, 4, 5),
  product_cd = c(1, 2, 3, 4, 5),
  qty        = c(10, 12, 9, 12, 8), 
  sales      = c(30, 60, 45, 24, 32)
  )

sales

# Dimension store
stores <- tibble(
  store_cd   = c(1, 2, 3, 4, 6),
  address    = c("1 main st", "20 side st", "19 square blvd", "101 first st", "1002 retail ave"),
  city       = c("Main", "East", "West", "North", "South"), 
  open_hours = c("7-23", "7-23", "9-21", "9-21", "9-21")
  )

stores

# Dimension product
products <- tibble(
  product_cd   = c(1, 2, 3, 4, 6),
  description  = c("Soft drink", "Frozen snack", "Fruit", "Water", "Fruit 2"),
  unit_price   = c(3.0, 5.0, 5.0, 2.0, 4.0), 
  unit_measure = c("each", "each", "kg", "each", "kg")
  )

products
```

## Left Join

Returns everything from the left table and only the matches from the right table.

If we want to know the description of the products, we can left join the sales table with the products table.

* Left table: _sales_
* Right table: _products_

```{r}
# Left join
sales %>% 
  left_join(products, by = "product_cd")
```


```{r}
# Left join with selected columns from products
sales %>%
  left_join(
    products %>%
      select(product_cd, description),
    by = "product_cd"
  )
```

## Right Join

Returns everything from the right table and only the matches from the left table.

If we want to join the sales number to the stores table, we can right join the stores table with the sales table.

* Left table: _sales_
* Right table: _stores_

```{r}
# Right join
sales %>%
  select(store_cd, sales) %>% 
  right_join(stores, by = "store_cd")
```

## Inner Join

Returns only the matching rows from the left and right tables.
When we inner join tables, the result will be only the observations present in each table.

* Left table: _sales_
* Right table: _stores_

```{r}
# Inner join
sales %>% 
  inner_join(stores, by = "store_cd")
```

## Full Join

Returns all the rows from both tables.

* Left table: _sales_
* Right table: _stores_

```{r}
# Full join
sales %>%
  full_join(stores)
```

## Anti-Join

Returns all the rows from the left table that are NOT in the right table.

* Left table: _sales_
* Right table: _products_

```{r}
# Anti-join
sales %>%
  anti_join(products)
```

# Do more with __{tidyverse}__

In this file, we will go over some interesting functions of the package __tidyverse__.

## Load a dataset
```{r}
data("mtcars")
```

## Functions from purrr

Functions similar to apply family

### Mapping functions

_map()_: applies function to every element of list or vector. Returns same data type.

```{r}
# Map
mtcars %>% 
  select(hp, wt) %>% 
  map(mean)
```

```{r}
# Map int
mtcars %>% 
  select(hp, wt) %>% 
  map_int(length)
```

## Binding Data

Binding data is useful to glue pieces of data together.

_brind rows_ glues the data by columns. So if your data has the same variables, you can use this function.

```{r}
# Creating datasets A and B
A <- mtcars %>% 
  slice_head(n = 3)

B <- mtcars %>% 
  slice_tail(n = 3)

# Bind rows
AB <- A %>%
  bind_rows(B) %>% 
  as_tibble(rownames = "names")

AB
```

_bind_cols()_ is to bind columns, gathering the data by rows. It is more suitable when the rows are for the same observation, but the columns are different and need to be put together.

```{r}
# Creating datasets A and B
A <- mtcars %>% 
  slice_head(n = 5) %>% 
  select(mpg:disp)

B <- mtcars %>% 
  slice_head(n = 5) %>% 
  select(hp:wt)

# Bind columns
AB <- A %>%
  bind_cols(B) %>% 
  as_tibble(rownames = "names")

AB
```

## Cumulative functions

Set of functions used along with _mutate()_, like _cumsum()_, _cumprod()_, _cummean()_ and _cummax()_. These are used for cumulative calculations of a variable using a simple syntax, requiring that we just pass a variable name to it within the mutate function, as seen next.

```{r}
# Cumulative sum of weight
mtcars %>%
  as_tibble(rownames = "names") %>% 
  mutate(cumulative_sum = cume_dist(wt)) %>%
  arrange(cumulative_sum)
```

## Case When

Function `case_when()` to deal with multiple cases of logical tests.

```{r}
# Case When to create a new label column for the transmission variable
mtcars %>%
  as_tibble(rownames = "names") %>%
  mutate(
    transmission_type = case_when(
      am == 0 ~ "automatic",
      am == 1 ~ "manual"
    )
  ) %>% 
  select(am, transmission_type)
```

## ggplot2

A basic plot with _ggplot2_

```{r}
# ggplot2 basic scatterplot
mtcars %>%
  as_tibble(rownames = "names") %>%
  ggplot(aes(hp, mpg)) +
  geom_point(color = "royalblue", size = 4, alpha = 0.5) +
  labs(title = "Relationship between HP vs. MPG") +
  theme_classic()
```
