---
title: "11 Enhanced Visualization with ggplot2"
author: "Gustavo R. Santos (original) | Antti Rask (modifications)"
date: "2022-08-26"
output: html_document
---

# Enhanced Visualization with {ggplot2}

## Import Libraries

```{r, message=FALSE, warning=FALSE}
library(conflicted)  # An Alternative Conflict Resolution Strategy 
library(datasets)    # The R Datasets Package
library(gghighlight) # Highlight Lines and Points in 'ggplot2'
library(ggrepel)     # Automatically Position Non-Overlapping Text Labels with 'ggplot2'     
library(patchwork)   # The Composer of Plots
library(plotly)      # Create Interactive Web Graphics via 'plotly.js'
library(tidyverse)   # Easily Install and Load the 'Tidyverse'
```

## Dataset

In this chapter, we will use the dataset "diamonds", from the package datasets  to illustrate the plots.

To make the code more generically applicable to any dataset, I will call it "tbl".

```{r}
# Loading the dataset
data("diamonds")

tbl <- diamonds

remove(diamonds)

tbl
```

A regular scatterplot of the prices by carat will look busy. Observe.

```{r}
# Scatterplot of price by carat.
tbl %>%
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  theme_classic()
```

```{r}
# Facet grid by cut of scatterplot price by carat
tbl %>%
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_grid(vars(cut)) +
  theme_bw()
```

```{r}
# Facet grid by cut and clarity of scatterplot price by carat
tbl %>% 
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_grid(vars(cut), vars(clarity)) +
  theme_bw()
```

__Facet wrap__

It forces the grid into a rectangular form.

```{r}
# Facet wrap by cut of scatterplot price by carat
tbl %>% 
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_wrap(vars(cut)) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
# Alternatively with {gghighlight}
tbl %>% 
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_wrap(vars(cut)) +
  theme_classic() +
  theme(legend.position = "none") +
  gghighlight()
```

### Map plots

Maps are good to plot information based on locations.

Let's load a file with location information about the USA states and their GDP in 2021.

https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_GDP

```{r}
# Load the CSV file
states <- read_csv("data/USA_states.csv")

states
```

Now, to create a map, we use the _geom_map()_ function.

```{r}
# Loading the map of the USA states from ggplot2
us <- map_data("state")

# Plot only the USA map
us %>% 
  ggplot() + 
  geom_map(
    aes(long, lat, map_id = region),
    map   = us,
    color = "black", 
    fill  = "lightgray"
  ) +
  theme_void()
```

Now let's plot the USA Map with the capital cities marked.

```{r}
# Names of the states to lower case
states_lower <- states %>%
  mutate(state = str_to_lower(state))

# Mapping the location of the state capitals
us_map <- states_lower %>% 
  ggplot() + 
  geom_map(
    aes(longitude, latitude, map_id = state),
    map   = us,
    color = "black",
    fill  = "lightgray"
  ) + 
  xlim(-130, -60) +
  ylim(20, 55) +
  theme_void()

options(scipen = 999)

# Adding the capital of the states locations.
us_map +
  geom_point(
    aes(longitude, latitude, size = GDP / 1000000, fill = GDP),
    shape = 24
  ) +
  labs(
    title    = "USA states and the capital cities",
    subtitle = str_glue(
    "The triangle is the location of the capital of the state.
    The bigger the triangle, the higher is the state GDP (USD Millions)"
    )
  )
```

We can also add the state names to the plot.

```{r}

# Mapping the location of the state capitals
states_lower %>% 
  ggplot() + 
  geom_map(
    aes(longitude, latitude, map_id = state),
    map   = us,
    color = "black",
    fill  = "lightgray"
  ) + 
  geom_point(
    aes(longitude, latitude),
    fill  = "coral",
    shape = 24
  ) +
  geom_text_repel(
    aes(longitude, latitude, label = state_cd),
    nudge_x = 2,
    color   = "darkred"
  ) +
  xlim(-130, -60) +
  ylim(20, 55) +
  theme_void()
```

### Time Series Plots

Time Series is a dataset where the sequence of points is ordered in time. So the data will present a measurement of any given variable through time periods, such as days, hours, months or any other time period.

Let's see how to visualize time series.

```{r}
set.seed(10)

# Creating a Dataset
ts <- tibble(
  date    = seq(ymd("2022-01-01"), ymd("2022-06-30"), by = "days"),
  measure = as.integer(runif(181, min = 600, max = 1000) + sort(rexp(181, 0.001)))
)

ts
```

Let's start with the most basic plot

```{r}
# Basic plot
ts %>%
  ggplot(aes(date, measure)) +
  geom_line(linewidth = 0.7) +
  theme_classic()
```

If we call the basic plot _basic_plot_, then all we have to do is to add more layers as we want.

```{r}
# Basic plot
basic_plot <- ts %>%
  ggplot(aes(date, measure)) +
  geom_line(linewidth = 0.7) +
  theme_classic()

# To change the X axis label format

# Year 4 digit, abbreviated month and day
basic_plot +
  scale_x_date(date_labels = "%Y %b %d")

# Abbreviated month, year 2 digit
basic_plot + 
  scale_x_date(date_labels = "%b %y")
```

We can also control the breaks between the labels. For example, we can see the breaks by month or bi-weekly instead of the three month period above.

```{r}
# Breaks by one month
basic_plot + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b")

# Breaks by bi-weekly month
basic_plot + 
  scale_x_date(date_breaks = "2 weeks", date_labels = "%W %y")

basic_plot + 
  scale_x_date(date_breaks = "2 weeks", date_labels = "%m %d")
```

We can also select some specific time dates from our time series.

```{r}
# Closeup on date after 06/01
basic_plot +
  scale_x_date(
    limit       = as_date(c("2022-06-01", "2022-07-01")),
    date_breaks = "1 week",
    date_labels = "%m %d"
  )
```

### 3D Plots

We can used 3D graphics to represent surfaces. To plot 3D, use the library __{plotly}__.

```{r}
# Set seet for reproduceability
set.seed(42)

# Creating a random surface
surface <- matrix(
  as.integer(
    sort(
      abs(
        runif(160, 90, 180)
        )
      )
    ),
  nrow = 80,
  ncol = 20
)

# Plotting the surface
plot_ly(z = ~surface) %>%
  add_surface()
```

Another good use of 3D graphics is to see the dimension where the separation of the points occur. Imagine you have tow variables points really close to each other and the separation happens on a third dimension.

That can be seen on a 3D plot.

```{r}
set.seed(42)

# Variables
var1 = rnorm(20, mean = 25, sd = 5)
var2 = var1 + 0.1

#data frame
my_data <- tibble(
  var1 = var1,
  var2 = var2,
  var3 = 1:20,
  var4 = rep(c("A","B"), each = 10)
)

# Plot 2D
my_data %>% 
  ggplot(aes(var1, var2, color = var4)) +
  geom_point() +
  theme_classic()
```

The 3D plot
```{r}
# Plot 3D
my_data %>%
  plot_ly(
    x      = ~var1,
    y      = ~var2,
    z      = ~var3,
    color  = ~var4,
    colors = c("royalblue", "coral")
  ) %>%
  add_markers()
```

```{r}
set.seed(42)

#data frame
my_data <- tibble(
  var1 = rnorm(2000, mean = 25, sd = 5),
  var2 = rnorm(2000),
  var3 = rnorm(2000)
)

# Plot 3D
my_data %>% 
  plot_ly(
    x      = ~var1,
    y      = ~var2,
    z      = ~var3,
    color  = ~var3,
    colors = c("royalblue", "coral", "yellow")
  ) %>% 
  add_markers()
```

### Adding interactivity to the graphics

Add interactivity is helpful because a graphic becomes  ready for a more in-depth analysis, being able to zoom in and out, for example.

To create an interactive plot out of ggplot2 graphic, just combine it with plotly function _ggplotly()_.

```{r}
# Interactive scatter plot

# Scatterplot of price by carat.
p <- tbl %>%
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  theme_classic()

ggplotly(p)
```

```{r}
#Interactive bar plot
p2 <- tbl %>% 
  ggplot(aes(cut)) + 
  geom_bar() +
  theme_classic()

ggplotly(p2)
```
