---
title: "Data Wrangling with R, Chapter 11: Enhanced Visualization with ggplot2"
author: "Gustavo R. Santos (original) | Antti Rask (modifications)"
date: "2022-08-26"
output: html_document
---

# Enhanced Visualization with {ggplot2}

## Load Libraries

```{r, message=FALSE, warning=FALSE}
library(conflicted)  # An Alternative Conflict Resolution Strategy
library(gghighlight) # Highlight Lines and Points in 'ggplot2'
library(ggrepel)     # Automatically Position Non-Overlapping Text Labels with 'ggplot2'     
library(patchwork)   # The Composer of Plots
library(plotly)      # Create Interactive Web Graphics via 'plotly.js'
library(tidyverse)   # Easily Install and Load the 'Tidyverse'
```

## Dataset

```{r}
# Loading the dataset
data("diamonds")

tbl <- diamonds

tbl
```

```{r}
# Scatterplot of price by carat
tbl %>%
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  theme_classic()
```

## Enhancements

### Faceting

#### Facet Grid

```{r}
# Facet grid by cut of scatterplot price by carat
tbl %>%
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_grid(vars(cut)) +
  theme_bw()
```

```{r}
# Facet grid by cut and clarity of scatterplot price by carat
tbl %>% 
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_grid(vars(cut), vars(clarity)) +
  theme_bw()
```

#### Facet wrap

It forces the grid into a rectangular form.

```{r}
# Facet wrap by cut of scatterplot price by carat
tbl %>% 
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_wrap(vars(cut)) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
# Alternatively with {gghighlight}
tbl %>% 
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  facet_wrap(vars(cut)) +
  theme_classic() +
  theme(legend.position = "none") +
  gghighlight()
```

### Map plots

Let's load a file with location information about the USA states and their GDP in 2021.

https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_GDP

```{r}
# Load the CSV file
states <- read_csv("data/USA_states.csv")

states
```

#### Plot the Map of USA

```{r}
# Loading the map of the USA states from ggplot2
us <- map_data("state")

# Plot only the USA map
us %>% 
  ggplot() + 
  geom_map(
    aes(long, lat, map_id = region),
    map   = us,
    color = "black", 
    fill  = "lightgray"
  ) +
  theme_void()
```

#### Map the Location of the State Capitals

```{r}
# Names of the states to lower case
states_lower <- states %>%
  mutate(state = str_to_lower(state))

# Mapping the location of the state capitals
us_map <- states_lower %>% 
  ggplot() + 
  geom_map(
    aes(longitude, latitude, map_id = state),
    map   = us,
    color = "black",
    fill  = "lightgray"
  ) + 
  xlim(-130, -60) +
  ylim(20, 55) +
  theme_void()

options(scipen = 999)

# Adding the capital of the states locations.
us_map +
  geom_point(
    aes(longitude, latitude, size = GDP / 1000000, fill = GDP),
    shape = 24
  ) +
  labs(
    title    = "USA states and the capital cities",
    subtitle = str_glue(
    "The triangle is the location of the capital of the state.
    The bigger the triangle, the higher is the state GDP (USD Millions)"
    )
  )
```

#### Map the Location of the State Capitals

```{r}
# Mapping the location of the state capitals
states_lower %>% 
  ggplot() + 
  geom_map(
    aes(longitude, latitude, map_id = state),
    map   = us,
    color = "black",
    fill  = "lightgray"
  ) + 
  geom_point(
    aes(longitude, latitude),
    fill  = "coral",
    shape = 24
  ) +
  geom_text_repel(
    aes(longitude, latitude, label = state_cd),
    nudge_x = 2,
    color   = "darkred"
  ) +
  xlim(-130, -60) +
  ylim(20, 55) +
  theme_void()
```

### Time Series Plots

```{r}
set.seed(10)

ts <- tibble(
  date    = seq(ymd("2022-01-01"), ymd("2022-06-30"), by = "days"),
  measure = as.integer(runif(181, min = 600, max = 1000) + sort(rexp(181, 0.001)))
)

ts
```

#### Basic plot

```{r}
ts %>%
  ggplot(aes(date, measure)) +
  geom_line(linewidth = 0.7) +
  theme_classic()
```

```{r}
# Basic plot
basic_plot <- ts %>%
  ggplot(aes(date, measure)) +
  geom_line(linewidth = 0.7) +
  theme_classic()

# To change the X axis label format

# Year 4 digit, abbreviated month and day
basic_plot +
  scale_x_date(date_labels = "%Y %b %d")

# Abbreviated month, year 2 digit
basic_plot + 
  scale_x_date(date_labels = "%b %y")
```

#### Date Breaks

```{r}
# Breaks by one month
basic_plot + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b")

# Breaks by bi-weekly month
basic_plot + 
  scale_x_date(date_breaks = "2 weeks", date_labels = "%W %y")

basic_plot + 
  scale_x_date(date_breaks = "2 weeks", date_labels = "%m %d")
```

#### Date Limits

```{r}
# Closeup on date after 06/01
basic_plot +
  scale_x_date(
    limit       = as_date(c("2022-06-01", "2022-07-01")),
    date_breaks = "1 week",
    date_labels = "%m %d"
  )
```

### 3D Plots

```{r}
set.seed(42)

# Creating a random surface
surface <- matrix(
  as.integer(
    sort(
      abs(
        runif(160, 90, 180)
        )
      )
    ),
  nrow = 80,
  ncol = 20
)

# Plotting the surface
plot_ly(z = ~surface) %>%
  add_surface()
```

```{r}
set.seed(42)

# Variables
var1 = rnorm(20, mean = 25, sd = 5)
var2 = var1 + 0.1

#data frame
my_data <- tibble(
  var1 = var1,
  var2 = var2,
  var3 = 1:20,
  var4 = rep(c("A","B"), each = 10)
)

# Plot 2D
my_data %>% 
  ggplot(aes(var1, var2, color = var4)) +
  geom_point() +
  theme_classic()
```

```{r}
# Plot 3D
my_data %>%
  plot_ly(
    x      = ~var1,
    y      = ~var2,
    z      = ~var3,
    color  = ~var4,
    colors = c("royalblue", "coral")
  ) %>%
  add_markers()
```

```{r}
set.seed(42)

#data frame
my_data <- tibble(
  var1 = rnorm(2000, mean = 25, sd = 5),
  var2 = rnorm(2000),
  var3 = rnorm(2000)
)

# Plot 3D
my_data %>% 
  plot_ly(
    x      = ~var1,
    y      = ~var2,
    z      = ~var3,
    color  = ~var3,
    colors = c("royalblue", "coral", "yellow")
  ) %>% 
  add_markers()
```

### Interactive Plots

```{r}
# Scatterplot of price by carat.
p <- tbl %>%
  ggplot(aes(carat, price, color = cut)) + 
  geom_point(alpha = 0.5) +
  theme_classic()

ggplotly(p)
```

```{r}
#Interactive bar plot
p2 <- tbl %>% 
  ggplot(aes(cut)) + 
  geom_bar() +
  theme_classic()

ggplotly(p2)
```
